---
output: github_document
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "man/figures/README-",
  out.width = "100%"
)
```

# robstore

<!-- badges: start -->
<!-- badges: end -->

R bindings to the Rust [`object_store`](https://crates.io/crates/object_store)
crate, providing a uniform interface to local filesystems and cloud object
stores. Built with [extendr](https://extendr.rs/).

The same small API (`store_put`, `store_get`, `store_get_range`, `store_list`,
`store_exists`, `store_delete`, `store_copy`) works across every backend —
switching between in-memory, local disk, AWS S3, S3-compatible endpoints
like Pawsey or source.coop, Google Cloud Storage, and Azure Blob Storage is
just a change of constructor.

For cloud workloads, `store_get_many()` and `store_get_ranges_many()` fan
requests out concurrently through a shared tokio runtime, delivering
10–30× speedups over sequential per-request calls.

## Installation

You will need a working Rust toolchain (see
[rustup.rs](https://rustup.rs/)). Then:

``` r
# install.packages("remotes")
remotes::install_github("mdsumner/robstore")
```

## Backends

| Constructor                                        | Use for                                                                |
|----------------------------------------------------|------------------------------------------------------------------------|
| `memory_store()`                                   | in-memory (tests, scratch)                                             |
| `local_store(path)`                                | local filesystem rooted at `path`                                      |
| `s3_store(...)`                                    | AWS S3 or S3-compatible with credentials                               |
| `s3_store_anonymous(...)`                          | public S3 buckets (e.g. `sentinel-cogs`, `source.coop`)                |
| `gcs_store(bucket)`                                | Google Cloud Storage with credentials (GOOGLE_APPLICATION_CREDENTIALS) |
| `gcs_store_anonymous(bucket)`                      | public GCS buckets (e.g. `gcp-public-data-arco-era5`)                  |
| `azure_store(account, container)`                  | Azure Blob Storage with env-var credentials                            |
| `azure_store_sas(account, container, sas_token)`   | Azure with a SAS token (e.g. Microsoft Planetary Computer)             |
| `azure_store_anonymous(account, container)`        | Azure unsigned (limited — no anonymous listing)                        |

Generic HTTP is planned.

## A quick tour

### In-memory

```{r memory, eval = FALSE}
library(robstore)

s <- memory_store()
s
#> <MemoryStore>

store_put(s, "hello.txt", charToRaw("Hello, world!"))
rawToChar(store_get(s, "hello.txt"))
#> [1] "Hello, world!"

store_list(s, prefix = NULL)
#> [1] "hello.txt"

store_exists(s, "hello.txt")
#> [1] TRUE

# byte-range read
store_get_range(s, "hello.txt", offset = 7, length = 5) |> rawToChar()
#> [1] "world"

store_copy(s, "hello.txt", "bye.txt")
store_delete(s, "hello.txt")
store_list(s, prefix = NULL)
#> [1] "bye.txt"
```

### Local filesystem

```{r local, eval = FALSE}
tmp <- tempfile("rs-"); dir.create(tmp)
ls <- local_store(tmp)
ls
#> <LocalStore(/tmp/rs-abc123)>

# nested keys work — intermediate directories are created automatically
store_put(ls, "a/b/c.txt", charToRaw("nested"))
list.files(tmp, recursive = TRUE)
#> [1] "a/b/c.txt"

rawToChar(store_get(ls, "a/b/c.txt"))
#> [1] "nested"
```

### Public S3 (anonymous) — Sentinel-2 COGs

A Sentinel-2 L2A scene from the public Element 84 archive, listed
anonymously:

```{r s3-anon, eval = FALSE}
s <- s3_store_anonymous(
  bucket    = "sentinel-cogs",
  region    = "us-west-2",
  endpoint  = NULL,
  allow_http = FALSE
)
s
#> <S3Store[anon](sentinel-cogs @ us-west-2)>

keys <- store_list(s, prefix = "sentinel-s2-l2a-cogs/1/C/CV/2024/1/")
length(keys)
#> [1] 63
head(keys, 3)
#> [1] "sentinel-s2-l2a-cogs/1/C/CV/2024/1/S2B_1CCV_20240106_0_L2A/AOT.tif"
#> [2] "sentinel-s2-l2a-cogs/1/C/CV/2024/1/S2B_1CCV_20240106_0_L2A/B01.tif"
#> [3] "sentinel-s2-l2a-cogs/1/C/CV/2024/1/S2B_1CCV_20240106_0_L2A/B02.tif"
```

### S3-compatible with credentials — Pawsey

For credentialed S3-compatible services (Pawsey, MinIO, Backblaze etc.),
point the `endpoint` argument at the provider and supply credentials via
the usual AWS environment variables:

```{r pawsey, eval = FALSE}
# credentials only needed for writes or private buckets
Sys.setenv(
  AWS_ACCESS_KEY_ID     = "<your-key-id>",
  AWS_SECRET_ACCESS_KEY = "<your-secret>"
)

s <- s3_store(
  bucket     = "estinel",
  region     = "",                                 # unused when endpoint is set
  endpoint   = "https://projects.pawsey.org.au",
  allow_http = FALSE
)

store_list(s, prefix = "sentinel-2-c1-l2a/2015")
#> [1] "sentinel-2-c1-l2a/2015/10/23/Hobart_2015-10-23.tif"
#> [2] "sentinel-2-c1-l2a/2015/11/12/Hobart_2015-11-12.tif"
#> [3] "sentinel-2-c1-l2a/2015/11/22/Hobart_2015-11-22.tif"
#> ...
```

### Parallel listing across prefixes

S3's `ListObjectsV2` is paginated (1000 keys per page) and each page
depends on the previous one's continuation token, so a single large
listing is serial by protocol. For hierarchical layouts — years, MGRS
tiles, product types — you can fan out across sub-prefixes with
`store_list_many()` and turn one long serial listing into many short
parallel ones.

Pawsey's `estinel` bucket is public, so `s3_store_anonymous()` works
against it too — no credentials needed:

```{r list-many, eval = FALSE}
s <- s3_store_anonymous(
  bucket     = "estinel",
  region     = "",
  endpoint   = "https://projects.pawsey.org.au",
  allow_http = FALSE
)

# serial — one paginated listing through the whole bucket
system.time({
  all_serial <- store_list(s, prefix = "sentinel-2-c1-l2a/")
})
#>    user  system elapsed
#>   2.441   0.657  43.132
length(all_serial)
#> [1] 552860

# parallel — one list call per year, fanned out
year_prefixes <- sprintf("sentinel-2-c1-l2a/%d/", 2015:2026)
system.time({
  all_parallel <- store_list_many(s, year_prefixes, concurrency = 12)
})
#>    user  system elapsed
#>   2.821   0.651   8.469
length(all_parallel)
#> [1] 552860
```

A 5× speedup over the serial listing, with the same 552,860 keys
returned (`store_list_many()` does not preserve input order — sort the
result if order matters).

### Multi-tenant buckets — source.coop

[`source.coop`](https://source.coop) is a multi-tenant S3 gateway: one
large public bucket (`us-west-2.opendata.source.coop`) with each data
provider getting a top-level prefix. From robstore's perspective it's
just an anonymous AWS S3 bucket — the bucket name has dots in it, but
that's the only unusual thing:

```{r source-coop, eval = FALSE}
s <- s3_store_anonymous(
  bucket     = "us-west-2.opendata.source.coop",
  region     = "us-west-2",
  endpoint   = NULL,
  allow_http = FALSE
)
s
#> <S3Store[anon](us-west-2.opendata.source.coop @ us-west-2)>

store_list_delimited(s, "ausantarctic/ghrsst-mur-v2/")
#> $keys
#> [1] "ausantarctic/ghrsst-mur-v2/README.md"
#> [2] "ausantarctic/ghrsst-mur-v2/ghrsst-mur-v2.parquet"
#>
#> $common_prefixes
#>  [1] "ausantarctic/ghrsst-mur-v2/2002"
#>  [2] "ausantarctic/ghrsst-mur-v2/2003"
#>  ...
#> [25] "ausantarctic/ghrsst-mur-v2/2026"

# fan-out list across years — same pattern as Pawsey
year_prefixes <- sprintf("ausantarctic/ghrsst-mur-v2/%d/", 2002:2026)
system.time({
  all_keys <- store_list_many(s, year_prefixes, concurrency = 16)
})
#>    user  system elapsed
#>   0.297   0.175   2.266
length(all_keys)
#> [1] 46070

# filter to the TIFs
tif_keys <- grepv("\\.tif$", all_keys)
tail(tif_keys)
#> [1] "ausantarctic/ghrsst-mur-v2/2026/04/15/20260415090000-JPL-L4_GHRSST-SSTfnd-MUR-GLOB-v02.0-fv04.1_sst_anomaly.tif"
#> ...
```

## Concurrent byte-range reads

`store_get_range()` is fast — reqwest over rustls with connection pooling —
but calling it in an R loop is still one request at a time.
`store_get_ranges_many()` issues up to `concurrency` range-GET requests
simultaneously through a shared tokio runtime, returning a list of raw
vectors in input order. This is the primitive for COG IFD walks,
Zarr chunk scatter-reads, and Parquet row-group reads.

```{r fanout, eval = FALSE}
s <- s3_store_anonymous("sentinel-cogs", "us-west-2", NULL, FALSE)
keys <- store_list(s, prefix = "sentinel-s2-l2a-cogs/1/C/CV/2024/1/")
length(keys)
#> [1] 63

# sequential — one GET at a time
system.time({
  heads_seq <- lapply(keys, \(k) store_get_range(s, k, 0, 65536))
})
#>    user  system elapsed
#>   0.026   0.065  19.333

# concurrent fan-out — 16 inflight
system.time({
  heads <- store_get_ranges_many(
    s,
    keys    = keys,
    offsets = rep(0, length(keys)),
    lengths = rep(65536, length(keys)),
    concurrency = 16
  )
})
#>    user  system elapsed
#>   0.036   0.044   1.625

# pushing concurrency higher
system.time({
  heads_32 <- store_get_ranges_many(
    s, keys,
    rep(0, length(keys)), rep(65536, length(keys)),
    concurrency = 32
  )
})
#>    user  system elapsed
#>   0.033   0.052   0.767
```

Note the `user`/`system` columns: essentially zero CPU work in R while Rust
drives the reactor.

### `store_head_bytes()` — convenience wrapper

For the common pattern of reading the first N bytes of many files (COG
headers, Parquet footers, Zarr `.zarray` files), `store_head_bytes()`
wraps `store_get_ranges_many()` with fixed offset and length:

```{r head, eval = FALSE}
hdrs <- store_head_bytes(s, keys, length = 65536, concurrency = 32)
length(hdrs)
#> [1] 63

# pass directly to your COG parser / IFD scanner of choice
```

### Scaling

At the time of writing, `store_list()` can return half a million keys
from a single prefix call (tested against `sentinel-s2-l2a-cogs/1/C/`
which returned 543,889 keys). For concurrent byte-range reads at that
scale, batch the keys in R rather than holding all the returned raw
vectors in memory at once:

```{r batched, eval = FALSE}
keys_huge <- store_list(s, prefix = "sentinel-s2-l2a-cogs/1/C/")
length(keys_huge)
#> [1] 543889

# process in chunks of 1000
batches <- split(keys_huge, ceiling(seq_along(keys_huge) / 1000))
for (batch in batches) {
  hdrs <- store_head_bytes(s, batch, length = 16384, concurrency = 64)
  # ... extract IFD offsets, write references, etc.
}
```

### Public GCS — Landsat and ERA5

`gcs_store_anonymous()` opens a public Google Cloud Storage bucket
with unsigned requests. Combined with `store_list_delimited()` it's
an efficient way to explore hierarchical archives like the Landsat
mirror:

```{r gcs-landsat, eval = FALSE}
s <- gcs_store_anonymous("gcp-public-data-landsat")

store_list_delimited(s, NULL)
#> $keys
#> [1] "index.csv.gz"
#>
#> $common_prefixes
#>  [1] "LC08" "LE07" "LM01" "LM02" "LM03" "LM04" "LM05" "LO08" "LT04" "LT05" "LT08"

store_list_delimited(s, "LC08/01/")$common_prefixes |> length()
#> [1] 233    # WRS-2 path directories

store_list_delimited(s, "LC08/01/090/")$common_prefixes |> length()
#> [1] 67     # rows within path 090
```

For analysis-ready Zarr stores, the cloud-native pattern is to fetch
the consolidated metadata and compute chunk keys rather than listing
them. The ERA5 archive on GCS exposes `.zmetadata` as a single ~130 KB
JSON file that describes the entire store:

```{r gcs-era5, eval = FALSE}
era <- gcs_store_anonymous("gcp-public-data-arco-era5")

meta <- store_get(era, "ar/full_37-1h-0p25deg-chunk-1.zarr-v3/.zmetadata")
length(meta)
#> [1] 132785

cat(substring(rawToChar(meta), 1, 400))
#> {"metadata": {".zattrs": {"valid_time_start": "1940-01-01",
#>  "last_updated": "2026-04-17 02:54:09...",
#>  "valid_time_stop": "2025-12-31", ...},
#>  "100m_u_component_of_wind/.zarray": {
#>    "chunks": [1, 721, 1440],
#>    "compressor": {"cname": "lz4", "id": "blosc", ...},
#>    "dtype": "<f4",
#>    "shape": [1323648, 721, 1440], ...}}
```

That 130 KB JSON describes 85 years of hourly global reanalysis across
hundreds of variables. The typical workflow is:

1. `store_get()` the consolidated metadata (one request).
2. Parse the JSON in R to learn shape, chunk grid, compressor, dtype.
3. Compute exactly the chunk keys you need for your `(variable, time,
   lat_range, lon_range)` window.
4. `store_get_many()` those chunks concurrently.

No listing of chunks — the metadata tells you the keys directly.
`robstore` provides the concurrent byte-fetch primitives; a Zarr-aware
layer (e.g. a future `zaro` integration) handles the codec pipeline
and array assembly.

### Azure Blob Storage — Microsoft Planetary Computer

Azure's auth model differs from S3 and GCS: anonymous listing is
generally not permitted even on "public" containers. Most public
Azure datasets — the Microsoft Planetary Computer collections in
particular — are accessed via short-lived SAS tokens from a free
token-minting endpoint:

```
https://planetarycomputer.microsoft.com/api/sas/v1/token/{account}/{container}
```

Fetch a token, hand it to `azure_store_sas()`, and listing and reads
work the same as any other backend:

```{r azure-sas, eval = FALSE}
library(httr2)

# Free SAS token from Planetary Computer (~1 hour lifetime)
token <- request(
  "https://planetarycomputer.microsoft.com/api/sas/v1/token/sentinel1euwestrtc/sentinel1-grd-rtc"
) |>
  req_perform() |>
  resp_body_json()

s <- azure_store_sas(
  account   = "sentinel1euwestrtc",
  container = "sentinel1-grd-rtc",
  sas_token = token$token
)
s
#> <AzureStore[sas](sentinel1euwestrtc/sentinel1-grd-rtc)>

# Navigate the Sentinel-1 GRD layout with delimited listings
store_list_delimited(s, NULL)$common_prefixes
#> [1] "GRD"

store_list_delimited(s, "GRD/")$common_prefixes
#>  [1] "GRD/2014" "GRD/2015" "GRD/2016" "GRD/2017" "GRD/2018" "GRD/2019"
#>  [7] "GRD/2020" "GRD/2021" "GRD/2022" "GRD/2023" "GRD/2024" "GRD/2025"
#> [13] "GRD/2026"

store_list_delimited(s, "GRD/2024/")$common_prefixes
#>  [1] "GRD/2024/1"  "GRD/2024/10" "GRD/2024/11" "GRD/2024/12"
#>  [5] "GRD/2024/2"  "GRD/2024/3"  ...
```

For your own Azure storage accounts, set the usual environment
variables and use `azure_store()`:

```{r azure-env, eval = FALSE}
Sys.setenv(
  AZURE_STORAGE_ACCOUNT_NAME = "myaccount",
  AZURE_STORAGE_ACCOUNT_KEY  = "..."
)
s <- azure_store("myaccount", "mycontainer")
```

`azure_store_anonymous()` exists for completeness but will fail on
most containers — Azure does not generally permit unsigned listing
on public data.

## API

All operations take a `Store` object as their first argument.

| Function                                                                | Description                                             |
|-------------------------------------------------------------------------|---------------------------------------------------------|
| `store_put(store, key, data)`                                           | write a raw vector                                      |
| `store_get(store, key)`                                                 | read an object as raw                                   |
| `store_get_range(store, key, offset, length)`                           | read a byte range                                       |
| `store_exists(store, key)`                                              | `TRUE`/`FALSE`                                          |
| `store_list(store, prefix)`                                             | character vector of keys                                |
| `store_copy(store, from, to)`                                           | copy within the same store                              |
| `store_delete(store, key)`                                              | remove an object                                        |
| `store_get_many(store, keys, concurrency)`                              | concurrent full-object reads                            |
| `store_get_ranges_many(store, keys, offsets, lengths, concurrency)`     | concurrent byte-range reads                             |
| `store_head_bytes(store, keys, length, offset, concurrency)`            | convenience wrapper for fixed-range reads               |
| `store_list_many(store, prefixes, concurrency)`                         | concurrent listings across many prefixes                |
| `store_list_delimited(store, prefix)`                                   | one-level hierarchical listing (keys + common prefixes) |

Keys are strings; paths like `"a/b/c.txt"` are handled the same way across
every backend.

## Development notes

Local, in-memory, AWS S3, S3-compatible, GCS, and Azure backends are
working and tested against:

- AWS — `sentinel-cogs` (anonymous)
- Pawsey — `estinel` (credentialed and anonymous)
- source.coop — `us-west-2.opendata.source.coop/ausantarctic/*`
- GCS — `gcp-public-data-arco-era5`, `gcp-public-data-landsat`
- Azure — Microsoft Planetary Computer via SAS tokens

Planned:

- generic HTTP backend
- vendored Rust dependencies (`rextendr::vendor_pkgs()`) for CRAN / r-universe
- integration with downstream packages (`rustycogs` for COG byte-range scanning,
  `zaro` for Zarr stores)

## Related

- [`object_store`](https://crates.io/crates/object_store) — the underlying
  Rust crate (Apache Arrow project)
- [`obstore`](https://github.com/developmentseed/obstore) — Python bindings
  to the same crate (Development Seed)
- [`extendr`](https://extendr.rs/) — the R↔Rust bridge used here

## License

MIT