---
output: github_document
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "man/figures/README-",
  out.width = "100%"
)
```

# robstore

<!-- badges: start -->
<!-- badges: end -->

R bindings to the Rust [`object_store`](https://crates.io/crates/object_store)
crate, providing a uniform interface to local filesystems and cloud object
stores. Built with [extendr](https://extendr.rs/).

The same small API (`store_put`, `store_get`, `store_get_range`, `store_list`,
`store_exists`, `store_delete`, `store_copy`) works across every backend —
switching between in-memory, local disk, AWS S3, and S3-compatible endpoints
like Pawsey is just a change of constructor.

## Installation

You will need a working Rust toolchain (see
[rustup.rs](https://rustup.rs/)). Then:

``` r
# install.packages("remotes")
remotes::install_github("mdsumner/robstore")
```

## Backends

| Constructor             | Use for                                          |
|-------------------------|--------------------------------------------------|
| `memory_store()`        | in-memory (tests, scratch)                       |
| `local_store(path)`     | local filesystem rooted at `path`                |
| `s3_store(...)`         | AWS S3 or S3-compatible with credentials         |
| `s3_store_anonymous(...)` | public S3 buckets (no signing, e.g. `sentinel-cogs`) |

More backends (GCS, Azure, generic HTTP) are planned.

## A tour of the API

### In-memory

```{r memory, eval = FALSE}
library(robstore)

s <- memory_store()
s
#> <MemoryStore>

store_put(s, "hello.txt", charToRaw("Hello, world!"))
rawToChar(store_get(s, "hello.txt"))
#> [1] "Hello, world!"

store_list(s, prefix = NULL)
#> [1] "hello.txt"

store_exists(s, "hello.txt")
#> [1] TRUE

# byte-range read
store_get_range(s, "hello.txt", offset = 7, length = 5) |> rawToChar()
#> [1] "world"

store_copy(s, "hello.txt", "bye.txt")
store_delete(s, "hello.txt")
store_list(s, prefix = NULL)
#> [1] "bye.txt"
```

### Local filesystem

```{r local, eval = FALSE}
tmp <- tempfile("rs-"); dir.create(tmp)
ls <- local_store(tmp)
ls
#> <LocalStore(/tmp/rs-abc123)>

# nested keys work — intermediate directories are created automatically
store_put(ls, "a/b/c.txt", charToRaw("nested"))
list.files(tmp, recursive = TRUE)
#> [1] "a/b/c.txt"

rawToChar(store_get(ls, "a/b/c.txt"))
#> [1] "nested"
```

### Public S3 (anonymous) — Sentinel-2 COGs

A Sentinel-2 L2A scene from the public Element 84 archive, listed
anonymously:

```{r s3-anon, eval = FALSE}
s <- s3_store_anonymous(
  bucket    = "sentinel-cogs",
  region    = "us-west-2",
  endpoint  = NULL,
  allow_http = FALSE
)
s
#> <S3Store[anon](sentinel-cogs @ us-west-2)>

keys <- store_list(s, prefix = "sentinel-s2-l2a-cogs/1/C/CV/2024/1/")
length(keys)
#> [1] 63
head(keys, 3)
#> [1] "sentinel-s2-l2a-cogs/1/C/CV/2024/1/S2B_1CCV_20240106_0_L2A/AOT.tif"
#> [2] "sentinel-s2-l2a-cogs/1/C/CV/2024/1/S2B_1CCV_20240106_0_L2A/B01.tif"
#> [3] "sentinel-s2-l2a-cogs/1/C/CV/2024/1/S2B_1CCV_20240106_0_L2A/B02.tif"
```

Byte-range reads work the same as any other store — useful for reading
COG headers without downloading the whole file:

```{r s3-range, eval = FALSE}
hdr <- store_get_range(s, keys[1], offset = 0, length = 65536)
length(hdr)
#> [1] 65536
```

### S3-compatible with credentials — Pawsey

Credentials are picked up from the standard `AWS_ACCESS_KEY_ID` /
`AWS_SECRET_ACCESS_KEY` environment variables. The `endpoint` argument
points `object_store` at any S3-compatible service:

```{r pawsey, eval = FALSE}
Sys.setenv(
  AWS_ACCESS_KEY_ID     = <key-id>,
  AWS_SECRET_ACCESS_KEY = <secret>)
)

s <- s3_store(
  bucket    = "estinel",
  region    = "",     
  endpoint  = "https://projects.pawsey.org.au",
  allow_http = FALSE
)

store_list(s, prefix = "sentinel-2-c1-l2a/2015")
#> [1] "sentinel-2-c1-l2a/2015/10/23/Hobart_2015-10-23.tif"
#> [2] "sentinel-2-c1-l2a/2015/11/12/Hobart_2015-11-12.tif"
#> [3] "sentinel-2-c1-l2a/2015/11/22/Hobart_2015-11-22.tif"
#> ...
```

## API

All operations take a `Store` object as their first argument.

| Function                                     | Description                          |
|----------------------------------------------|--------------------------------------|
| `store_put(store, key, data)`                | write a raw vector                   |
| `store_get(store, key)`                      | read an object as raw                |
| `store_get_range(store, key, offset, length)`| read a byte range                    |
| `store_exists(store, key)`                   | `TRUE`/`FALSE`                       |
| `store_list(store, prefix)`                  | character vector of keys             |
| `store_copy(store, from, to)`                | copy within the same store           |
| `store_delete(store, key)`                   | remove an object                     |

Keys are strings; paths like `"a/b/c.txt"` are handled the same way
across every backend.

## Listing large buckets

`store_list()` returns *all* keys matching the prefix. Public buckets like
`sentinel-cogs` contain millions of objects, so always pass a sufficiently
narrow `prefix` — or be prepared to wait and consume memory. The prefix is
evaluated server-side for cloud backends, so a narrow prefix is cheap.

## Development notes

Phase 1 (local/in-memory) and Phase 2 (S3 + S3-compatible) are working and
tested against `sentinel-cogs` (AWS) and Pawsey. Planned:

- GCS, Azure, generic HTTP backends
- vendored Rust dependencies (`rextendr::vendor_pkgs()`) for CRAN / r-universe
- integration with downstream packages (`rustycogs` for COG byte-range scanning,
  `zaro` for Zarr stores)

## Related

- [`object_store`](https://crates.io/crates/object_store) — the underlying
  Rust crate (Apache Arrow project)
- [`obstore`](https://github.com/developmentseed/obstore) — Python bindings
  to the same crate
- [`extendr`](https://extendr.rs/) — the R↔Rust bridge used here

## License

MIT