---
output: github_document
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "man/figures/README-",
  out.width = "100%"
)
```

# robstore

<!-- badges: start -->
<!-- badges: end -->

R bindings to the Rust [`object_store`](https://crates.io/crates/object_store)
crate, providing a uniform interface to local filesystems and cloud object
stores. Built with [extendr](https://extendr.rs/).

The same small API (`store_put`, `store_get`, `store_get_range`, `store_list`,
`store_exists`, `store_delete`, `store_copy`) works across every backend —
switching between in-memory, local disk, AWS S3, and S3-compatible endpoints
like Pawsey is just a change of constructor.

For cloud workloads, `store_get_many()` and `store_get_ranges_many()` fan
requests out concurrently through a shared tokio runtime, delivering
10–30× speedups over sequential per-request calls.

## Installation

You will need a working Rust toolchain (see
[rustup.rs](https://rustup.rs/)). Then:

``` r
# install.packages("remotes")
remotes::install_github("mdsumner/robstore")
```

## Backends

| Constructor               | Use for                                          |
|---------------------------|--------------------------------------------------|
| `memory_store()`          | in-memory (tests, scratch)                       |
| `local_store(path)`       | local filesystem rooted at `path`                |
| `s3_store(...)`           | AWS S3 or S3-compatible with credentials         |
| `s3_store_anonymous(...)` | public S3 buckets (no signing, e.g. `sentinel-cogs`) |
| `store_list_many(store, prefixes, concurrency)`   | concurrent listings across many prefixes |

More backends (GCS, Azure, generic HTTP) are planned.

## A tour of the API

### In-memory

```{r memory, eval = FALSE}
library(robstore)

s <- memory_store()
s
#> <MemoryStore>

store_put(s, "hello.txt", charToRaw("Hello, world!"))
rawToChar(store_get(s, "hello.txt"))
#> [1] "Hello, world!"

store_list(s, prefix = NULL)
#> [1] "hello.txt"

store_exists(s, "hello.txt")
#> [1] TRUE

# byte-range read
store_get_range(s, "hello.txt", offset = 7, length = 5) |> rawToChar()
#> [1] "world"

store_copy(s, "hello.txt", "bye.txt")
store_delete(s, "hello.txt")
store_list(s, prefix = NULL)
#> [1] "bye.txt"
```

### Local filesystem

```{r local, eval = FALSE}
tmp <- tempfile("rs-"); dir.create(tmp)
ls <- local_store(tmp)
ls
#> <LocalStore(/tmp/rs-abc123)>

# nested keys work — intermediate directories are created automatically
store_put(ls, "a/b/c.txt", charToRaw("nested"))
list.files(tmp, recursive = TRUE)
#> [1] "a/b/c.txt"

rawToChar(store_get(ls, "a/b/c.txt"))
#> [1] "nested"
```

### Public S3 (anonymous) — Sentinel-2 COGs

A Sentinel-2 L2A scene from the public Element 84 archive, listed
anonymously:

```{r s3-anon, eval = FALSE}
s <- s3_store_anonymous(
  bucket    = "sentinel-cogs",
  region    = "us-west-2",
  endpoint  = NULL,
  allow_http = FALSE
)
s
#> <S3Store[anon](sentinel-cogs @ us-west-2)>

keys <- store_list(s, prefix = "sentinel-s2-l2a-cogs/1/C/CV/2024/1/")
length(keys)
#> [1] 63
head(keys, 3)
#> [1] "sentinel-s2-l2a-cogs/1/C/CV/2024/1/S2B_1CCV_20240106_0_L2A/AOT.tif"
#> [2] "sentinel-s2-l2a-cogs/1/C/CV/2024/1/S2B_1CCV_20240106_0_L2A/B01.tif"
#> [3] "sentinel-s2-l2a-cogs/1/C/CV/2024/1/S2B_1CCV_20240106_0_L2A/B02.tif"
```

### S3-compatible with credentials — Pawsey

Credentials are picked up from the standard `AWS_ACCESS_KEY_ID` /
`AWS_SECRET_ACCESS_KEY` environment variables. The `endpoint` argument
points `object_store` at any S3-compatible service:

```{r pawsey, eval = FALSE}
Sys.setenv(
  AWS_ACCESS_KEY_ID     = <key-id>,
  AWS_SECRET_ACCESS_KEY =<secret>
)

s <- s3_store(
  bucket    = "estinel",
  region    = "",                                  # unused when endpoint is set
  endpoint  = "https://projects.pawsey.org.au",
  allow_http = FALSE
)

store_list(s, prefix = "sentinel-2-c1-l2a/2015")
#> [1] "sentinel-2-c1-l2a/2015/10/23/Hobart_2015-10-23.tif"
#> [2] "sentinel-2-c1-l2a/2015/11/12/Hobart_2015-11-12.tif"
#> [3] "sentinel-2-c1-l2a/2015/11/22/Hobart_2015-11-22.tif"
#> ...
```

### Parallel listing across prefixes

S3's `ListObjectsV2` is paginated (1000 keys per page) and each page
depends on the previous one's continuation token, so a single large
listing is serial by protocol. For hierarchical layouts — years, MGRS
tiles, product types — you can fan out across sub-prefixes with
`store_list_many()` and turn one long serial listing into many short
parallel ones.

```{r list-many, eval = FALSE}
s <- s3_store(
  bucket    = "estinel",
  region    = "",
  endpoint  = "https://projects.pawsey.org.au",
  allow_http = FALSE
)

# serial — one paginated listing through the whole bucket
system.time({
  all_serial <- store_list(s, prefix = "sentinel-2-c1-l2a/")
})
#>    user  system elapsed
#>   2.441   0.657  43.132
length(all_serial)
#> [1] 552860

# parallel — one list call per year, fanned out
year_prefixes <- sprintf("sentinel-2-c1-l2a/%d/", 2015:2026)
system.time({
  all_parallel <- store_list_many(s, year_prefixes, concurrency = 12)
})
#>    user  system elapsed
#>   2.821   0.651   8.469
length(all_parallel)
#> [1] 552860
```

A 5× speedup over the serial listing, with the same 552,860 keys
returned (`store_list_many()` does not preserve input order — sort the
result if order matters).

## Concurrent byte-range reads

`store_get_range()` is fast — reqwest over rustls with connection pooling —
but calling it in an R loop is still one request at a time.
`store_get_ranges_many()` issues up to `concurrency` range-GET requests
simultaneously through a shared tokio runtime, returning a list of raw
vectors in input order. This is the primitive for COG IFD walks,
Zarr chunk scatter-reads, and Parquet row-group reads.

```{r fanout, eval = FALSE}
s <- s3_store_anonymous("sentinel-cogs", "us-west-2", NULL, FALSE)
keys <- store_list(s, prefix = "sentinel-s2-l2a-cogs/1/C/CV/2024/1/")
length(keys)
#> [1] 63

# sequential — one GET at a time
system.time({
  heads_seq <- lapply(keys, \(k) store_get_range(s, k, 0, 65536))
})
#>    user  system elapsed
#>   0.026   0.065  19.333

# concurrent fan-out — 16 inflight
system.time({
  heads <- store_get_ranges_many(
    s,
    keys    = keys,
    offsets = rep(0, length(keys)),
    lengths = rep(65536, length(keys)),
    concurrency = 16
  )
})
#>    user  system elapsed
#>   0.036   0.044   1.625

# pushing concurrency higher
system.time({
  heads_32 <- store_get_ranges_many(
    s, keys,
    rep(0, length(keys)), rep(65536, length(keys)),
    concurrency = 32
  )
})
#>    user  system elapsed
#>   0.033   0.052   0.767
```

Note the `user`/`system` columns: essentially zero CPU work in R while Rust
drives the reactor.

### `store_head_bytes()` — convenience wrapper

For the common pattern of reading the first N bytes of many files (COG
headers, Parquet footers, Zarr .zarray files), `store_head_bytes()` wraps
`store_get_ranges_many()` with fixed offset and length:

```{r head, eval = FALSE}
hdrs <- store_head_bytes(s, keys, length = 65536, concurrency = 32)
length(hdrs)
#> [1] 63

# pass directly to your COG parser / IFD scanner of choice
```

### Scaling

At the time of writing, `store_list()` can return half a million keys
from a single prefix call (tested against `sentinel-s2-l2a-cogs/1/C/`
which returned 543,889 keys). For concurrent byte-range reads at that
scale, batch the keys in R rather than holding all the returned raw
vectors in memory at once:

```{r batched, eval = FALSE}
keys_huge <- store_list(s, prefix = "sentinel-s2-l2a-cogs/1/C/")
length(keys_huge)
#> [1] 543889

# process in chunks of 1000
batches <- split(keys_huge, ceiling(seq_along(keys_huge) / 1000))
for (batch in batches) {
  hdrs <- store_head_bytes(s, batch, length = 16384, concurrency = 64)
  # ... extract IFD offsets, write references, etc.
}
```

## API

All operations take a `Store` object as their first argument.

| Function                                          | Description                          |
|---------------------------------------------------|--------------------------------------|
| `store_put(store, key, data)`                     | write a raw vector                   |
| `store_get(store, key)`                           | read an object as raw                |
| `store_get_range(store, key, offset, length)`     | read a byte range                    |
| `store_exists(store, key)`                        | `TRUE`/`FALSE`                       |
| `store_list(store, prefix)`                       | character vector of keys             |
| `store_copy(store, from, to)`                     | copy within the same store           |
| `store_delete(store, key)`                        | remove an object                     |
| `store_get_many(store, keys, concurrency)`        | concurrent full-object reads         |
| `store_get_ranges_many(store, keys, offsets, lengths, concurrency)` | concurrent byte-range reads |
| `store_head_bytes(store, keys, length, offset, concurrency)` | convenience wrapper for fixed-range reads |

Keys are strings; paths like `"a/b/c.txt"` are handled the same way across
every backend.

## Development notes

Phase 1 (local/in-memory), Phase 2 (S3 + S3-compatible), and Phase 3
(concurrent fan-out) are working and tested against `sentinel-cogs` (AWS)
and Pawsey. Planned:

- GCS, Azure, generic HTTP backends
- vendored Rust dependencies (`rextendr::vendor_pkgs()`) for CRAN / r-universe
- integration with downstream packages (`rustycogs` for COG byte-range scanning,
  `zaro` for Zarr stores)

## Related

- [`object_store`](https://crates.io/crates/object_store) — the underlying
  Rust crate (Apache Arrow project)
- [`obstore`](https://github.com/developmentseed/obstore) — Python bindings
  to the same crate
- [`extendr`](https://extendr.rs/) — the R↔Rust bridge used here

## License

MIT